# Reproducible Research: Peer Assessment 1
```{r}
library(lattice)
library(ggplot2)
library(mice)
library(plyr)
library(knitr)
library(grid)
library(gridExtra)
```
## Loading and preprocessing the data
```{r}
data<-read.csv("activity/activity.csv")
data[,"date"]<-as.Date(data$date)
```
## Count NA
```{r}
sum(is.na(data))   
```
## What is mean total number of steps taken per day?
```{r}
totalStepsPerDay<-ddply(data,.(date),summarize,Total=sum(steps,na.rm=TRUE))
```

## Create Histogram
## Determine Number of Bins
```{r}
class = nclass.Sturges(totalStepsPerDay$Total)
```
## Calculate bin width
```{r}
bin=range(totalStepsPerDay$Total)/class
```
## Plot Histogram 
```{r}
ggplot(totalStepsPerDay,aes(x=Total))+geom_histogram(aes(fill = ..count..),binwidth = bin[2])
```

## What is mean total number of steps taken per day?
## mean
```{r}
mean(totalStepsPerDay$Total)
```
## median
```{r}
median(totalStepsPerDay$Total)
```


## What is the average daily activity pattern?
```{r}
averageStepsPerinterval<-ddply(data,.(interval),summarize,meanSteps=mean(steps,na.rm=TRUE))
```
## max mean Steps interval
```{r}
averageStepsPerinterval[which.max(averageStepsPerinterval$meanSteps),]
```
## Plot Mean Steps vs interval
```{r}
ggplot(averageStepsPerinterval, aes(x=interval, y=meanSteps))+geom_line(aes(colour = meanSteps),size=1)+scale_colour_gradient(high="#56B4E9",low="#D55E00")
```

## Imputing missing values
Add an extra column where date column is converted to a factor and then a numeric. I did this to make imputing with mice package
work with date as a variable.  Results seem reasonable. However, more diagnostics should probably be done.
```{r}
timeData = data.frame(data,time=as.numeric(factor(data$date)))
miceData = data.frame(steps=timeData$steps,date=timeData$time,interval=timeData$interval)
```

```{r cache=TRUE}
imp <- imp <- mice(miceData, m = 5, seed = 23109)
```
```{r}
stripplot(imp, pch = 20, cex = 1.2)
```

```{r}
xyplot(imp, steps ~ interval | .imp, pch = 20, cex = 1.4)
```

```{r}
imputedData=complete(imp)
recombinedData = data.frame(steps=imputedData$steps,date=data$date,interval=imputedData$interval)
```
## What is mean total number of steps taken per day?
```{r}
imputedTotalStepsPerDay<-ddply(recombinedData,.(date),summarize,Total=sum(steps))
```

## Create Histogram
## Determine Number of Bins
```{r}
class = nclass.Sturges(imputedTotalStepsPerDay$Total)
```
## Calculate binWidth
```{r}
bin=range(imputedTotalStepsPerDay$Total)/class
```
## Plot Histogram 
```{r}
ggplot(imputedTotalStepsPerDay,aes(x=Total))+geom_histogram(aes(fill = ..count..),binwidth = bin[2])
```

## What is mean total number of steps taken per day?
## mean
```{r}
mean(imputedTotalStepsPerDay$Total)
```
## median
```{r}
median(imputedTotalStepsPerDay$Total)
```
## Are there differences in activity patterns between weekdays and weekends?
```{r}
recombinedData<-transform(recombinedData ,week = ifelse(weekdays(data$date) %in% c('Sunday','Saturday') ,"Weekend","Weekday" ))
```
```{r}
newaverageStepsPerinterval<-ddply(recombinedData,.(interval,week),summarize,meanSteps=mean(steps))
```
```{r}
ggplot(averageStepsPerinterval, aes(x=interval, y=meanSteps))+geom_line(aes(colour = meanSteps),size=1)+scale_colour_gradient(high="#56B4E9",low="#D55E00")
```
```{r}
weekday=ggplot(newaverageStepsPerinterval, aes(x=interval, y=meanSteps))+geom_line(aes(colour = meanSteps),size=1,subset=.(week == 'Weekday'))+scale_colour_gradient(high="#56B4E9",low="#D55E00")
```
```{r}
weekend=ggplot(newaverageStepsPerinterval, aes(x=interval, y=meanSteps))+geom_line(aes(colour = meanSteps),size=1,subset=.(week == 'Weekend'))+scale_colour_gradient(high="#56B4E9",low="#D55E00")
```
```{r}
grid.arrange(weekday,weekend, ncol=1)
```