# Reproducible Research: Peer Assessment 1
```{r}
library(lattice)
library(ggplot2)
library(mice)
library(plyr)
library(knitr)
library(grid)
library(gridExtra)
```
## Loading and preprocessing the data
```{r}
data<-read.csv("activity/activity.csv")
data[,"date"]<-as.Date(data$date)
```
### Count NA
```{r}
sum(is.na(data))   
```
## What is mean total number of steps taken per day?
### Create Histogram

Total steps per day was calculated using ddply.
```{r}
totalStepsPerDay<-ddply(data,.(date),summarize,Total=sum(steps,na.rm=TRUE))
```
#### Determine Number of Bins
```{r}
class = nclass.Sturges(totalStepsPerDay$Total)
```
#### Calculate bin width
```{r}
bin=range(totalStepsPerDay$Total)/class
```
#### Plot Histogram 
```{r}
ggplot(totalStepsPerDay,aes(x=Total))+geom_histogram(aes(fill = ..count..),binwidth = bin[2])
```

#### Mean
```{r}
mean(totalStepsPerDay$Total)
```
#### Median
```{r}
median(totalStepsPerDay$Total)
```


## What is the average daily activity pattern?

Average steps in an interval was first calculated with ddply.
```{r}
averageStepsPerinterval<-ddply(data,.(interval),summarize,meanSteps=mean(steps,na.rm=TRUE))
```
#### Max mean Steps in an interval
```{r}
averageStepsPerinterval[which.max(averageStepsPerinterval$meanSteps),]
```
#### Plot Mean Steps vs Interval
```{r}
ggplot(averageStepsPerinterval, aes(x=interval, y=meanSteps))+geom_line(aes(colour = meanSteps),size=1)+scale_colour_gradient(high="#56B4E9",low="#D55E00")
```

## Imputing missing values
Data was imputed with multiple imputation using the mice package. The method used was predictive mean matching. An extra column was added where the date column is converted to a factor and then a numeric. I did this to make imputing with mice package work. Results seem reasonable. However, more diagnostics should probably be done. Extra iterations should also probably be done to decrese the noise. Only 5 iterations were run.
```{r}
timeData = data.frame(data,time=as.numeric(factor(data$date)))
miceData = data.frame(steps=timeData$steps,date=timeData$time,interval=timeData$interval)
```

```{r cache=TRUE}
imp <- imp <- mice(miceData, m = 5, seed = 23109)
```
```{r}
stripplot(imp, pch = 20, cex = 1.2)
```

```{r}
xyplot(imp, steps ~ interval | .imp, pch = 20, cex = 1.4)
```

```{r}
imputedData=complete(imp)
recombinedData = data.frame(steps=imputedData$steps,date=data$date,interval=imputedData$interval)
```
## What is mean total number of steps taken per day?
```{r}
imputedTotalStepsPerDay<-ddply(recombinedData,.(date),summarize,Total=sum(steps))
```

### Create Histogram
#### Determine Number of Bins
```{r}
class = nclass.Sturges(imputedTotalStepsPerDay$Total)
```
#### Calculate bin width
```{r}
bin=range(imputedTotalStepsPerDay$Total)/class
```
#### Plot Histogram 
```{r}
ggplot(imputedTotalStepsPerDay,aes(x=Total))+geom_histogram(aes(fill = ..count..),binwidth = bin[2])
```

#### Mean
```{r}
mean(imputedTotalStepsPerDay$Total)
```
#### Median
```{r}
median(imputedTotalStepsPerDay$Total)
```
## Are there differences in activity patterns between weekdays and weekends?
```{r}
recombinedData<-transform(recombinedData ,week = ifelse(weekdays(data$date) %in% c('Sunday','Saturday') ,"Weekend","Weekday" ))
```

#### Mean Steps in an interval with Imputed Data
```{r}
newaverageStepsPerinterval<-ddply(recombinedData,.(interval,week),summarize,meanSteps=mean(steps))
```

```{r}
ggplot(newaverageStepsPerinterval, aes(x=interval, y=meanSteps))+geom_line(aes(colour = meanSteps),size=1)+scale_colour_gradient(high="#56B4E9",low="#D55E00")
```
#### Comparing Weekend and Weeday Patterns with Imputed Data
```{r}
weekday=ggplot(newaverageStepsPerinterval, aes(x=interval, y=meanSteps))+geom_line(aes(colour = meanSteps),size=1,subset=.(week == 'Weekday'))+scale_colour_gradient(high="#56B4E9",low="#D55E00")+ ggtitle("Weekday")
```
```{r}
weekend=ggplot(newaverageStepsPerinterval, aes(x=interval, y=meanSteps))+geom_line(aes(colour = meanSteps),size=1,subset=.(week == 'Weekend'))+scale_colour_gradient(high="#56B4E9",low="#D55E00")+ ggtitle("Weekend")
```
```{r}
grid.arrange(weekday,weekend, ncol=1)
```